---
title: "GitHub R Intro"
author: "Tess Noble Strohm"
date: "2023-01-19"
output: html_document
---

```{r set up packages}
source("setup.R")
```

Load and See the Data

```{r Load and see data}
data("penguins")
penguins
class(penguins)
str(penguins)
class(penguins$species)
str(penguins$species)
```

Indexing

```{r Indexing}
myList <- list("apple", 1993, FALSE, penguins)
str(myList)

# for lists we use double brackes [[]]
myList[[1]]
myList[["data"]]

# for vectors we use single brackets []
myVector <- c("apple", "banana", "pear")
myVector[2]

# dataframe[row(s), columns()]
penguins[1:5, 2]

penguins[1:5, "island"]

penguins[1, 1:5]

penguins[1:5, c("species","sex")]

penguins[penguins$sex=='female',]

# $ for a single column
penguins$species
```

**6.3.1 Exercises**

```{r}
##1
myList["Fruit"] #Does not work because this is a list, therefore you need DOUBLE brackets, also 'fruit' is not on the list, not column name
myList[["Fruit"]]

penguins$flipper_lenght_mm #"length' is misspelled
penguins$flipper_length_mm

penguins[island=='Dream',] #need to specify the dataset AGAIN before telling it what column to look at, comma is NEEDED to end the arguement 
penguins[penguins$island=='Dream',]

##2 How many species in the data set, what islands are the data from?
unique(penguins$species)
unique(penguins$island)
summary(penguins) #3 species, 3 islands (Biscoe, Dream, Torgersen)

##3 Use indexing to create a new data frame that has only 3 columns: species, island and flipper length columns, and subset all rows for just the ‘Dream’ island.
penguins[penguins$island == "Dream", c("island", "species", "flipper_length_mm")] #to do this: info about ROWS goes on the left, columns on the right 

##4 Use indexing and the mean() function to find the average flipper length for the Adelie species on Dream island. (Note: explore the mean() function and how to deal with NA values).
subset <- penguins[penguins$island == "Dream" & penguins$species == "Adelie",]
mean(subset$flipper_length_mm)
```

**6.4 The dplyr package**

```{r}
filter(penguins, species == "Adelie")

filter(penguins, species != "Adelie")

filter(penguins, island %in% c("Dream", "Torgersen") & !is.na(bill_length_mm))
```

using the select() function

```{r}
# Select two specific variables
select(penguins, species, sex)

# Select a range of variables
select(penguins, species:flipper_length_mm)

# Rename columns within select
select(penguins, genus = species, island)

# Select column variables that are recorded in mm
select(penguins, contains("mm"))
```

Create new variables with mutate()

```{r}
# New variable that calculates bill length in cm
mutate(penguins, bill_length_cm = bill_length_mm/10)

# mutate based on conditional statements
mutate(penguins, species_sex = if_else(sex == 'male', paste0(species,"_m"), paste0(species, "_f")))
```

**6.4.1 The pipe %\>%**
Ex: Say you want to summarize data by some specified group, for example you want to find the average body mass for each species, this is where the group_by() function comes into play.
```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(body_mass_avg = mean(body_mass_g, na.rm = TRUE))
```

**6.3.2 Exercises**
Reorder the variables in penguins so that year is the first column followed by the rest (Hint: look into the use of everything()).
```{r}
penguins %>% select(year, everything())
```

Create a new column called ‘size_group’ where individuals with body mass greater than the overall average are called ‘large’ and those smaller are called ‘small’.
```{r}
penguins %>% 
  mutate(size_group=if_else(body_mass_g > mean(body_mass_g, na.rm = TRUE), "large", "small"))
```

Find out which year for each species were individuals on average the largest according to body mass.
```{r}
penguins %>%
  group_by(species) %>%
  filter(body_mass_g == max(body_mass_g, na.rm = TRUE))
```

You want to filter data for years that are not in a vector of given years, but this code doesn’t work. Tweak it so that it does. (Yes, you could just filter year to equal 2007 in this case but there is a trouble-shooting lessons here).
```{r}
penguins %>% 
  filter(year !%in% c(2008, 2009))

penguins %>% 
  filter(!(year %in% c(2008, 2009)))
```

**7. Visualize**
If you plan on doing any statistical analysis on your data , one of the first things you are likely to do is explore the distribution of your variables. You can plot histograms with geom_histogram()
```{r}
ggplot(penguins) + 
  geom_histogram(mapping = aes(x = flipper_length_mm))
```

This tells us there may be a lot of variation in flipper size among species. We can use the ‘fill =’ argument to color the bars by species, and scale_fill_manual() to specify the colors.
```{r}
# Histogram example: flipper length by species
ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm, fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4"))
```
Cool, now we can see there seems to be some pretty clear variation in flipper size among species. Another way to visualize across groups is with facet_wrap(), which will create a separate plot for each group, in this case species.
```{r}
ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm, fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~species)
```
Lets make a quick bar plot showing the total count of each species studied on each island
```{r}
ggplot(penguins) +
  geom_bar(mapping = aes(x = island, fill = species))
```
As you may have already noticed, the beauty about ggplot2 is there are a million ways you can customize your plots. This example builds on our simple bar plot
```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"), 
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```
We can use geom_point() to view the relationship between two variables by specifying the x and y axes. Say we want to visualize the relationship between penguin body mass and flipper length and color the point by species:
```{r}
ggplot(penguins) +
  geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm, color = species))
```

**7.1 Exercises**
Make a barplot showing the average flipper length for each species.
```{r}
ggplot(penguins) + geom_bar(mapping = aes(x=flipper_length_mm, fill=species))
```

Make a scatter plot of bill length compared to bill depth but only for observations on the Dream island.
```{r}
penguins %>%
  filter(island == "Dream") %>%
  ggplot() +
  geom_point(mapping = aes(x=bill_length_mm, y=bill_depth_mm))

penguins %>%
  filter(island == "Dream")
```




```{r}
#import colorado counties with tigris
counties <- counties(state = "CO")

#import roads for Larimer County
roads <- roads(state = "CO", county = "Larimer")

#set tmap road to interactive
tmap_mode("view")

qtm(counties)
tm_shape(counties)+
  tm_polygons()
```
